---
title: "Group_10_Analysis"
format: 
 html: 
  embed-resources: true
editor: visual
execute: 
 echo: false
 eval: true
 warning: false
 error: false
 message: false
---

# Data Analysis Presentation

#Introduction
 Measles is a highly contagious disease, and is a significant risk to the public, especially for children that are left unvaccinated. In 1998, the release of the Wakefield Article, which had detailed a link between the MMR vaccine and an increased risk of autism, was released. This in turn resulted in a drop in vaccination rates to roughly 80% in 2003 over the UK. The article was somewhat withdrawn in 2004, after studies found the two were falsely linked, and in fact were not associated.  
This study investigates the measles susceptibility of school children across Edinburgh's 101 intermediate zones (IZ’s) between 1998 and 2012. It aims to discover whether the retraction of the Wakefield article had any effect on Edinburgh's measle susceptibility, and if so, determine if these changes occurred in 2004, coinciding with the article retraction. This will hopefully indicate the importance of accuracy in article information, and why it is crucial that medical articles are reliable, so that overall public health is not at risk. 

# Exploratory Data Analysis {#EDA}

## Question 1

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false

library(tidyverse)
library(tidymodels)
library(moderndive)
library(skimr)
library(magrittr)
library(gapminder)
library(gridExtra)
library(gt)
```

```{r}
#| echo: false

dataset<- read.csv("dataset.csv")

prop<- dataset$Y/dataset$N

dataset$prop <- prop
```

```{r}
#| echo: false
#| eval: true
#| fig-cap: Scatterplot of Proportion of susceptible children to measles from 1998 to 2012
#| label: fig-1

ggplot(data = dataset, 
       mapping = aes(x = Year , y = prop))+ 
         labs(x="Year", y="Proportion", title="Scatterplot of Proportion of susceptible children to measles over the period 1998 to 2012") +
  geom_point()


```

-   @fig-1 shows a slight decline in proportion. However there seems to be a lot of overlapping points so we will jitter it to get a better look at the relationship.

```{r}
#| echo: false
#| eval: true
#| fig-cap: Scatterplot (jittered) of Proportion of susceptible children to measles over the period 1998 to 2012
#| label: fig-2



ggplot(dataset, aes(x=Year, y=prop))+
  geom_jitter()+
  labs(x="Year", y="Proportion", title="Scatterplot of Proportion of susceptible children to measles over the period 1998 to 2012") 


```

-   Plotting the jittered scatter plot @fig-2 shows a clearer decline in proportion of susceptible children from 1998 to 2012, but it should be noted that there is missing data in 2006 so this may make the graph harder to interpret. In addition we can see that towards the end of the time period there is less variance in the proportion of susceptible children, and we will investigate this further.

```{r}
#| echo: false
#| eval: true

skim(dataset)


```

-   We used the skim function to have a look at the histograms for the different variables. So we can see that the number of children in each IZ is normally distributed, but if we look at the histogram for the Y variable, number of children susceptible to measles in a given IZ, is not normally distributed which would suggest that there is another variable which is impacting the number of susceptible children. This supports the question of interest that there was a change in children's susceptibility to measles overtime.

```{r}
#| echo: false
#| eval: true

cor<- dataset|>
  get_correlation(formula= prop~Year)

gt(cor)

```

-   We calculated the correlation coefficient. It produced a value of -0.3553176. Whilst this is a weaker result, it is not insignificant, and does suggest there is a decreasing relationship. Again, this supports the idea that susceptibility has decreased over the years.

```{r}
#| echo: false
#| eval: true
#| fig-cap: Boxplot of Proportion of susceptible children to measles over the period 1998 to 2012
#| label: fig-3

ggplot(dataset, aes(x=Year, y=prop, group=Year))+
  geom_boxplot()+
    labs(x="Year", y="Proportion", title="Boxplot of Proportion of susceptible children to measles over the period 1998 to 2012") 

```

-   So we can see also from the @fig-3 that the proportion of susceptible children is decreasing over the years. In particular, we can see that the mean is decreasing slowly overtime, and in the later years it shows that there is less variation in the number of susceptible children, therefore it is becoming more steady and consistent.

## Question 2

```{r}
#| echo: false
#| eval: true

dataset_before<- subset(dataset, dataset$Year==1998 | dataset$Year== 2000 | dataset$Year==2002)

dataset_after<- subset(dataset, dataset$Year==2004 | dataset$Year== 2006 | dataset$Year==2008 | dataset$Year==2010 | dataset$Year==2012)




```

-   So we can analyse the difference before and after the articles retraction in 2004, we have split the dataset into two smaller ones, one containing the data before, and one containing the data after the retraction.

```{r}
#| echo: false
#| eval: true


cor_before<- dataset_before|>
  get_correlation(formula= prop~Year)

cor_after<- dataset_after|>
  get_correlation(formula= prop~Year)
```

| Correlation before | Correlation after |
|--------------------|-------------------|
| 0.01648565         | -0.3870859        |

: Correlations before and after the retraction

-   We calculated the correlation coefficient for the data before and after the retraction. We can very clearly see from the correlation coefficients that after the retraction that the was a significant change in relationship, as before there seemed to be no relationship as the value was 0.01648565, compared to after where the value was -0.3870859, which is clear difference, showing a decreasing relationship.

```{r}
#| echo: false
#| eval: true
#| fig-cap: Scatterplot (jittered) of Proportion of susceptible children to measles over the period 1998 to 2004
#| label: fig-4

ggplot(data = dataset_before, 
       mapping = aes(x = Year , y = prop)) + 
  geom_jitter()



```

```{r}
#| echo: false
#| eval: true
#| fig-cap: Scatterplot (jittered) of Proportion of susceptible children to measles over the period 2004 to 2012
#| label: fig-5
ggplot(data = dataset_after, mapping = aes(x = Year , y = prop)) + geom_jitter()
```

-   Finally, we can see that in the data before the retraction @fig-4 there is not much a trend between year and proportion, it seems fairly constant. When we compare this to the graph for after the retraction @fig-5 we can see a trend, which is a decreasing trend.

```{r}
#| echo: false
#| eval: true
#| fig-cap: Boxplot of Proportion of susceptible children to measles over the period 1998 to 2004, and 2004 to 2012
#| label: fig-6

boxplot_before<- ggplot(dataset_before, aes(y = prop)) +
  geom_boxplot() +
  labs( y = "Prop", 
       title = "before")

boxplot_after<- ggplot(dataset_after, aes(y = prop)) +
  geom_boxplot() +
  labs(y = "Prop", 
       title = "after")

grid.arrange(boxplot_before, boxplot_after, ncol = 2, nrow = 1)


```

-   We can see that the mean is lower after the retraction of the article in 2004,, than the mean before the retraction, before 2004, @fig-6. In addition we can see that the range of data is narrower after the retraction showing that the results were less varied and more consistent. This would also suggest that the proportion of children that are susceptible is becoming more steady and consistent.

```{r}
#| echo: false
#| eval: true
#| fig-cap: Scatterplot (jittered) with line of best fit, of Proportion of susceptible children to measles over the period 1998 to 2004, and 2004 to 2012
#| label: fig-7

jitplot_before<- ggplot(dataset_before, aes(x = Year, y = prop)) +
  geom_jitter()+
  geom_smooth(method = "lm", se = FALSE)

jitplot_after<- ggplot(dataset_after, aes(x = Year, y = prop)) +
  geom_jitter()+
  geom_smooth(method = "lm", se = FALSE)


grid.arrange(jitplot_before, jitplot_after, ncol = 2, nrow = 1)

```

-   Before 2004 we can see there is no significant trend between year and the proportion of children susceptible, wheras after this we can see a decreasing linear trend between year and proportion of children susceptible to measles, @fig-7.


# Formative Data Analysis

```{r}
setwd("C:\\Users\\Theo\\OneDrive - University of Glasgow\\Desktop\\DA Group data")
dataset <- read.csv("DAProject16.csv")
library(tidymodels)
library(parsnip)
library(magrittr)
library(dplyr)
library(ggplot2)

dataset <- dataset |> mutate(proportions = Y/N)
dataset$Post2004 <- ifelse(dataset$Year >= 2004, 1, 0)

```

We must model our data to include an indicator variable examining the difference between the number of people susceptible to measles before and after the article retraction.

It's appropriate to model this data under a binomial assumption as long as we specify our successes/ failures appropriately:

We fit the GLM:

$$
g(p_i) = logit(p_i) = \alpha + \beta_{year} \cdot year + \beta_{post2004} \cdot \mathbb{I}_{post2004} + \beta_{year,post2004} \cdot year \cdot \mathbb{I}_{post2004}
$$

So that,

$$
g(p_i) = logit(p_i) = \begin{cases} 
\alpha + \beta_{year} \cdot year {\text{,  for years before 2004}} \\
\alpha + \beta_{year} \cdot year + \beta_{post2004} + \beta_{year,post2004} \cdot year {\text{,  otherwise}}
\end{cases}
$$

This is where:

-   $\alpha$ is our model intercept.

-   $\beta_{year}$ is our model coefficient for 'Year' (in terms of an increase/decrease in log-odds).

-   $\beta_{post2004}$ is our beta coefficient for the log-odds of susceptibility for years after 2004 compared to before.

-   $\mathbb{I}_{post2004}$ is our indicator function where:

$$
\mathbb{I}_{post2004} = \begin{cases}
1 ~ \text{if the observation is taken post-2004} \\
0 ~ \text{otherwise}
\end{cases}
$$

-   $\beta_{year,post2004}$ is our interaction term between our explanatory variables.

Here's a visualisation of our model:

```{r}

dataset$successes <- dataset$N - dataset$Y  

model_bin <- glm(cbind(Y, successes) ~ Year + Post2004 + Year:Post2004, 
             family = binomial, data = dataset)

summary(model_bin)
```

However, we must be careful to avoid modelling with overdispersion. The overdispersion ratio - $\hat{\phi} = \frac{\text{residual deviance}}{\text{df}_{\text{residual deviance}}}$ - gives us information on the overdispersion.

```{r}
overdispersion_ratio <- summary(model_bin)$deviance / summary(model_bin)$df.residual
print(overdispersion_ratio)
```

This value is far greater than 1. This suggests that we have heavy overdispersion and we should adjust our model.

```{r}
model_quasi <- glm(cbind(Y, successes) ~ Year * Post2004, 
                   family = quasibinomial, data = dataset)
summary(model_quasi)
```

Displayed above is our quasi-binomial model. The standard errors have changed to so that our intervals for significance are adjusted for overdispersion.

But, before we infer from our model, we have an extremely large coefficient for our 'Post2004' variable. It's likely that - because of the scale of our 'Year' variable - this caused some interpretability issues.

Taking years before 2004 to be negative and years post 2004 to be positive:

```{r}
dataset <- dataset |> mutate(year_centered = Year - 2004)
centered_model <- glm(cbind(Y, successes) ~ year_centered * Post2004, 
                        family = quasibinomial, data = dataset)
summary(centered_model)
```

Here our results are centered and easier to interpret.

To answer our first question: "Did Edinburgh exhibit a change in measles susceptibility following the retraction of the Wakefield article?" we examine our 'year_centered:Post2004' coefficient. Our coefficient is highly significant and negative, indicating a significant decrease in the log odds of susceptibility. Suggesting that there is actually a decrease in measles susceptibility after the retraction of the Wakefield article post 2004.

For our second question: "Did the change, if any, in measles susceptibility occur in 2004 alongside the articles’ retraction?" the answer appears to be no. Given our 'Post2004' variable isn't significant. We can deduce that there existed no immediate jump in susceptibility in measles after the retraction. The change progressed - long term - afterwards.

Pseudo $R^2$ gives us a measure of model fit.

$$
R^2 = 1 - \frac{\text{Residual deviance}}{\text{Null deviance}}
$$

```{r}
R2_McFadden <- 1 - (1283.6 / 1596.0)
print(R2_McFadden)

```

Therefore, our model explains 19.6% of variation in measles susceptibility. This is a weak to moderate fit. However, because of the complication within the modelling for our centered variable, indicator variable and our overdispersion this is a reasonable model to use.

Here are our diagnostic plots:

```{r}
library(performance)
check_model(centered_model, panel = TRUE)
```

As we can see - for our Binned Residuals check - most points like within our error bounds. Suggesting that our model predicted probabilities align fairly well with our observed data. For our influencial observaitons, all points lie within our region suggesting no major influencial outliers. We have some high collinearity, specifically for Year variable and our interaction. Suggesting that our uncertainty could be fairly high in our estimates. Finally, our residuals appear to be normally distributed.

Overall our assumptions seem quite reasonable for using this model on our data.

#Conclusion

We can see throughout the analyis that it highlights the sigificant impact spreading misinformation has on public health, and how vital it is for information to be correct, so the public don’t jump to conclusions and in doing so negatively affect their health. Due to the negative value for our coefficient, with p-value <0.05, we can see that there is a significant decrease in measles susceptibility, after the Wakefield Article was withdrawn. We can also see that due to the less significant coefficient, the change in measles susceptibility was a long-term one, and did not occur instantly in 2004. This is likely to be expected, as the public take time to adjust to new pieces of research, especially ones that go against previous beliefs that have been set. It is also worth noting that the research was only partially retracted in 2004, and not fully discredited until 2010, hence suggesting a more long-term change is likely. 
It is also worth noting the models pseudo-R^2 indicates a moderate fit of the data, and as such we could look to improve the model by considering other variables that could have a significant effect on the date, such as the location of the IZ. People are likely to follow their peers, and people act on decisions related to political/religious beliefs, and as these can vary quite significantly from location to location around a city, this could be a crucial variable that could overall influence the measles susceptibility across Edinburgh. 
